---
title: "Geocomputation with R"
author: "Rick Dean"
format: 
  html:
    toc: false
    toc-depth: 4
    toc-location: "right"
    number-sections: true
    number-offset: 2
    self-contained: true
    smooth-scroll: true
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    code-overflow: wrap
    tbl-cap-location: "bottom"
    fig-width: 7
    fig-asp: 0.6
    fig-align: "center"
    fig-cap-location: "bottom"
    css: ../../style.css
    minimal: false
    link-external-newwindow: true
    callout-appearance: simple
    callout-icon: false
    callout-border-left: "#22CC00"
    abstract-title: "Abstract"
    abstract: "Contained here are notes on [3 Attribute data operations](https://geocompr.robinlovelace.net/attr.html) from the book [Geocomputation with R](https://geocompr.robinlovelace.net/index.html) by Lovelace, Nowosad, Muenchow. We will stay close to the book's major sections with some modifications/additions in the R scripts based on this author's own preferences."
---

::: note
-   Article setup:
    -   Install \`quarto\` executable from \<https://quarto.org\>
-   To compile article inside RStudio:
    -   open file `3_Attribute_data_operations.qmd`
    -   press \`Render\` from a recent version of RStudio
-   For a table of contents, set the yaml line `toc:` to true and re-render.
:::

# Attribute data operations

::: task
Load the Required R Packages.
:::

```{r, message=FALSE, warning=FALSE}
#| label: load-packages

library(data.table, quietly = T)
library(sf, quietly = T)
library(terra, quietly = T)
library(spData, quietly = T)
library(magrittr, quietly = T)
library(ggplot2, quietly = T)
library(RplotterPkg, quietly = T)
library(RspatialPkg, quietly = T)
```

## Introduction

> Attribute data is non-spatial information associated with geographic (geometry) data.

## Vector attribute manipulation

> The geometry column of `sf` objects is typically called `geometry` or `geom` but any name can be used.

::: task
In the `spData::world` data.frame change the names of the `geom` and `name_long` columns to `g` and `n` respectively using base R data.frame techniques.
:::

Make a copy of `spData::world` and change the copies' column names:

```{r}
world_df <- cbind(spData::world)
colnames(world_df)[which(names(world_df) == "geom")] <- "g"
colnames(world_df)[which(names(world_df) == "name_long")] <- "n"
colnames(world_df)
```

With the geometry renamed to "g", set the geometry column of `world_df` and show it and data.frame's class:

```{r}
sf::st_geometry(world_df) <- "g"
class(world_df$g)
class(world_df)
```

### Vector attribute subsetting

::: task
Select rows of `world_df` for countries with a small area.
:::

Perform the logical filter via `data.table` techniques:

```{r}
world_small_sf <- data.table::as.data.table(spData::world) %>% 
  .[area_km2 < 40000,] %>% 
  sf::st_as_sf(.)
```

```{r}
#| code-fold: true
#| fig.cap: Map of countries whose area is less than 40000 km2.

world_small_plot <- RspatialPkg::get_geom_sf(
    sf = world_small_sf,
    sf_fill = "red",
    hide_x_tics = T,
    hide_y_tics = T
  )
world_small_plot
```

### Chaining commands with pipes

::: task
Use data.table chaining commands to filter, select columns and rows of `spData::world`.
:::

Create the simple feature:
Get the first 5 asian countries.
```{r}
world_asia_sf <- data.table::as.data.table(spData::world) %>%
  .[continent == "Asia"] %>%
  .[1:5, .(name_long, continent, geom)] %>%
  sf::st_as_sf(.)
```

```{r}
#| code-fold: true
#| fig-cap: Subset an sf object of countries using data.table selection.

world_asia_plot <- RspatialPkg::get_geom_sf(
  sf = world_asia_sf,
  aes_text = "name_long",
  hide_x_tics = T,
  hide_y_tics = T
)
world_asia_plot
```

### Vector attribute aggregation

::: task
Aggregate the `spData::world` sf object on the "continent" column to give population, area, and number of observations.
:::
Perform the chained aggregation:

```{r}
world_agg_chain_sf <- data.table::as.data.table(spData::world) %>%
  .[, 
    .(Pop = sum(pop, na.rm = TRUE), 
      Area = sum(area_km2), 
      N = .N, 
      Density = round(sum(pop, na.rm = T)/sum(area_km2)), 
      geometry = sf::st_union(geom)), by = continent] %>%
  .[order(-N)] %>%
  .[1:6,] %>%
  sf::st_as_sf(.)
```

```{r}
#| code-fold: true
#| fig-cap: Aggregation/grouping of a sf object using data.table techniques to compute a country's population density.

world_agg_chain_plot <- RspatialPkg::get_geom_sf(
    sf = world_agg_chain_sf,
    aes_fill = "Density",
    panel_color = "green",
    hide_x_tics = T,
    hide_y_tics = T,
    legend_key_width = 0.7,
    legend_key_height = 1.0
  ) +
  RspatialPkg::get_geom_sf(
    sf = world_agg_chain_sf,
    aes_text = "continent",
    text_color = "red",
    adding = T
  ) +
ggplot2::scale_fill_gradientn(
  colors = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"),
  n.breaks = 8
)  +
ggplot2::guides(
  fill = ggplot2::guide_colorbar(
    ticks.colour = "black"
  )
)
world_agg_chain_plot
```


Display the resulting attributes of the sf object in a table:

```{r}
#| code-fold: true
#| tbl-cap: Summary of country attributes from data.table grouping.

RplotterPkg::create_table(
  x = as.data.table(world_agg_chain_sf)[,1:5],
  container_width_px = 300
)
```

### Vector attribute joining

::: task
Using `data.table` techniques join data on coffee production (`spData::coffee_data`) with the `spData::world` dataset with the "name_long" column as the common variable.
:::

```{r}
coffee_data_dt <- data.table::as.data.table(spData::coffee_data)
world_coffee_sf <- data.table::as.data.table(spData::world) %>% 
  coffee_data_dt[., on = c("name_long", "name_long")] %>% 
  .[, .(name_long, coffee_production_2017, geom)] %>% 
  sf::st_as_sf(.)
world_coffee_sf
```

```{r}
#| code-fold: true
#| fig-cap: Coffee production around the world.

RspatialPkg::get_geom_sf(
  sf = world_coffee_sf,
  aes_fill = "coffee_production_2017",
  subtitle = "Coffee Production for 2017",
  center_titles = T,
  legend_key_width = 0.7,
  legend_key_height = 1.0
) +
ggplot2::scale_fill_gradientn(
  colors = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"),
  n.breaks = 8
)  +
ggplot2::guides(
  fill = ggplot2::guide_colorbar(
    ticks.colour = "black"
  )
)
```

### Creating attributes and removing spatial information

::: task
Join the character columns "continent" and "region_un" into one character column called "con_reg" from `spData::world`.
:::

```{r}
world_join_sf <- data.table::as.data.table(spData::world) %>% 
  .[, `:=`(con_reg = paste(continent, region_un, sep = ":"), continent = NULL, region_un = NULL)] %>% 
  sf::st_as_sf(.)
head(world_join_sf)
```

::: task
Split the character column of `world_join_sf$con_reg` back into separate columns of "continent" and "region_un".
:::

```{r}
world_split_sf <- as.data.table(world_join_sf) %>% 
  .[, c("continent", "region_un") := tstrsplit(con_reg, ":")] %>% 
  sf::st_as_sf(.)
head(world_split_sf)
```

::: task
Rename the "name_long" column of `spData::world` to "name".
:::

```{r}
world_rename_sf <- data.table::as.data.table(spData::world) %>% 
  data.table::setnames(., old = "name_long", new = "name") %>% 
  sf::st_as_sf(.)
colnames(world_rename_sf)
```

## Manipulating raster objects

::: task
Create from scratch a simple SpatRaster object with 6 rows and columns using `terra::rast()`.
:::

Create the dataset using `terra::rast()`:

```{r}
elev_scratch_sr <- terra::rast(
  nrows = 6,
  ncols = 6,
  resolution = 0.5,
  xmin = -1.5,
  xmax = 1.5,
  ymin = -1.5,
  ymax = 1.5,
  vals = 1:36,
  names = "scratch"
)
elev_scratch_sr
```

```{r}
#| code-fold: true
#| fig-cap: Mapping of a scratch `SpatRaster` created from `terra::rast()`.

terra::plot(elev_scratch_sr)
```

3.  Display the SpatRaster object using `RspatialPkg::get_geom_raster()`.

```{r}
#| code-fold: true
#| fig.cap: Map a scratch SpatRaster object using` RspatialPkg::get_geom_raster()`. 

elev_scratch_df <- as.data.frame(elev_scratch_sr, xy = TRUE)
RspatialPkg::get_geom_raster(
  df = elev_scratch_df,
  aes_x = "x",
  aes_y = "y",
  aes_fill = "scratch",
  legend_key_width = 0.7,
  legend_key_height = 1.0
) +
ggplot2::scale_fill_gradientn(
  colors = rev(grDevices::terrain.colors(50)),
  n.breaks = 8
) +
ggplot2::guides(
  fill = ggplot2::guide_colorbar(
    ticks.colour = "black"
  )
)
```

::: task
Create the same SpatRaster object from a dataframe.
:::

Create the data.frame:

```{r}
x <- seq(-1.25, 1.25, length.out = 6)
y <- seq(-1.25, 1.25, length.out = 6)
elev_scratch_df <- expand.grid(x = x, y = y)
elev_scratch_df$z <- 1:36
```

Convert the data.frame to a SpatRaster object:

```{r}
elev_scratch_df_sr <- terra::rast(elev_scratch_df)
terra::crs(elev_scratch_df_sr) <- "EPSG:4326"
```

```{r}
#| code-fold: true
#| fig-cap: Mapping of a scratch `SpatRaster` created from a  data.frame.

elev_scratch_df_df <- as.data.frame(elev_scratch_df_sr, xy = TRUE)
RspatialPkg::get_geom_raster(
  df = elev_scratch_df_df,
  aes_x = "x",
  aes_y = "y",
  aes_fill = "z",
  legend_key_width = 0.7,
  legend_key_height = 1.0
) +
ggplot2::scale_fill_gradientn(
  colors = rev(grDevices::terrain.colors(50)),
  n.breaks = 8
) +
ggplot2::guides(
  fill = ggplot2::guide_colorbar(
    ticks.colour = "black"
  )
)
```

::: task
Create a raster object from a factor variable.
:::

Create the factor variable:

```{r}
grain_order_v <- c("clay","silt","sand")
grain_values_v <- sample(grain_order_v, 36, replace = TRUE)
grain_factor <- factor(grain_values_v, levels = grain_order_v)
```

Create a raster object from the factor:

```{r}
grain_sr <- terra::rast(
  nrow = 6,
  ncols = 6,
  resolution = 0.5,
  xmin = -1.5,
  xmax = 1.5,
  ymin = -1.5,
  ymax = 1.5,
  vals = grain_factor
)
grain_sr
```

```{r}
#| code-fold: true
#| fig-cap: Map of a scratch `SpatRaster` object that is defined from a factor/categorial variable.

terra::plot(grain_sr)
```

### Raster subsetting

::: task
Retrieve the top left pixel in the terra raster object `elev_scratch_sr`.
:::

```{r}
cell_1_1_df <- elev_scratch_sr[1,1]
cell_1_1_df$scratch
```

::: task
Retrieve the values of all the cells of `grain_sr`.
:::

```{r}
grain_values_mat <- grain_sr[]
grain_values_mat[,1]
```

### Summarizing raster objects

::: task
Create a frequency table of `grain_sr` by calling `terra::freq()` function.
:::

```{r}
terra::freq(grain_sr)
```

### Conclusion
This concludes the notes on Chapter 3 "Attribute data operations".
